Longitudinal Principal Manifold Estimation

### 2024-04-04 paper/main.tex
rearrange figures
update adni application section with new results

While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, for the purposes of demonstration we used $T_1$-weighted structural MRI images from eight study participants in the cognitively normal group (\cite{jack2008adni}). Follow-up duration for these participants ranged between approximately 16 months to 48 months, with imaging scheduled to be conducted at six month intervals. Individuals from the cognitively normal group were selected because drastic changes in the volume and shape of subcortical structures over the course of the study follow-up duration would be unexpected. Under this assumption, we attribute volumetric changes in a given structure to error introduced by the study procedure or the image segmentation process. Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.

@@

The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data. To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously. Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.

To understand the impacts of these discrepancies on the estimated volume of the regions, the volume of the hippocampus was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point. The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume. The interior identification approach developed by \cite{mengPrincipalManifoldEstimation2021} is used to determine whether each voxel may be considered inside the surface of the structure or not. The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}. The volume estimates can be found in Figure \ref{fig:lhipp_volumes}. Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations.

The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data. To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same healthy individuals described previously. Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus. This is reflected in the smaller discrepancies between volume estimates, as seen in Figure \ref{fig:lthal_volumes}, though visible variations between time points remain.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
  \caption{Left Hippocampus Cross Section. We can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the surface in the observed data. In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it. Gaps in the surface estimates frequently correspond with inaccurate volume estimates.}
  \label{fig:lhipp_cross_sections}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section}
  \caption{Left Thalamus Cross Section. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
  \label{fig:lthal_cross_sections}
\end{figure}

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|c c c c|}
    \hline
    Structure & Data & LPME & PME  \\
    \hline
    Left Hippocampus & 218 & 225 & 191 \\
    Right Hippocampus & 221 & 245 & 195 \\
    Left Thalamus & 82.9 & 49.7 & 72.0 \\
    Right Thalamus & 147 & 116 & 130 \\
    \hline
  \end{tabular}
  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant.}
  \label{table:adni_volume_sds}
\end{table}

>>>>

While the original study included 200 cognitively healthy elderly individuals, 400 with mild cognitive impairment, and 200 with AD, this analysis focuses on 297 participants, of whom 85 were considered cognitively normal, 150 displayed symptoms of mild cognitive impairment, and 62 were diagnosed with AD at the baseline study visit (\cite{jack2008adni}). Of the 85 cognitively normal participants, 28 were subsequently diagnosed with mild cognitive impairment or dementia during the course of the study. Follow-up duration for these participants ranged from approximately 6 months to 112 months, with imaging scheduled to be conducted at six- or 12-month intervals depending on the stage of the study. To encourage greater stability of longitudinal estimates, only participants with at least 24 months of follow up were considered, meaning that the following analysis was conducted with data from 250 participants. Due to the changes experienced by those with Alzheimer's disease, we focused on the hippocampi of the selected individuals.

@@

The hippocampus is a region with a relatively irregular shape, which may account for the differences in performance between the LPME algorithm when applied to simulated data and when applied to the hippocampus data. To understand how the approach performs when used with a more regularly shaped brain region, we also applied the PME and LPME algorithms to the surface of the left thalamus of the same set of individuals described previously. Estimates of the surface of the thalamus were obtained from the MRI images using the same preprocessing and segmentation steps detailed above for the hippocampus data.

Considering the same individual shown in Figures \ref{fig:adni_result} and \ref{fig:adni_cross_section}, results of fitting the PME and LPME algorithms to the surface of the left thalamus are shown in Figures \ref{fig:adni_lthal_result} and \ref{fig:lthal_cross_sections}. While the segmentation estimates of the thalamus do not show the same sort of deviations that are clearly visible for the hippocampus, the cross section plots shown in Figure \ref{fig:lthal_cross_sections} indicate that both the PME and LPME algorithms show a closer fit to the more regularly shaped thalamus surface than was achieved for the hippocampus.

To understand how the subcortical structure estimates obtained by the LPME and PME algorithms may impact the corresponding estimates of the volumes of these structures, the volume of each structure was approximated from the raw data and the estimates obtained by the PME and LPME algorithms at each time point. The volume estimation approach used here relies on counting the voxels contained within the boundary defined by the PME- or LPME-estimated embedding map, with each voxel having a known volume.  The code used to implement this approach is available at \texttt{https://github.com/rjzielinski/lpme-project}. It is important to note that, instead of using the volume estimates produced by FSL's FIRST method during the segmentation process, the same volume estimation method used for the LPME- and PME-based estimates is used to reach estimates based on the raw data as well to align processing steps as closely as possible.

The volume estimates for randomly selected individuals can be found in Figures \ref{fig:lhipp_volume_comparison} and \ref{fig:lthal_volume_comparison}. Here, we see that volume estimates differ substantially depending on whether PME, LPME, or the raw data were used, with the LPME estimate resulting in the lowest volume computations. In Figure \ref{fig:lhipp_volume_comparison}, we see that the LPME-based volume estimates for the hippocampus are consistently lower than those obtained from the raw data and the PME estimates. All three sources show substantial variability between study visits, particularly for subjects 137\_S\_0283 and 137\_S\_0686. In Figure \ref{fig:lthal_volume_comparison}, we again see that meaningful variability in the volume estimates using the raw data and PME estimates as sources exist between time points. However, the LPME estimates demonstrate the ability to smooth over this variability, resulting in volume estimates that are not as sensitive to differences between time points. We also see that when basing volume estimates on the more reliable LPME fit to the thalamus data, LPME yields volume estimates similar in magnitude to those obtained from the raw data and PME estimates.

A summary of the between-visit variability in the volume estimates of the subcortical structures considered here is shown in Table \ref{table:adni_volume_sds}. In cases without regression adjustment, we estimated the standard deviation of the volume estimates for each individual and each estimate source (i.e. raw data, LPME, and PME). We then took the mean of these values for each estimate source. When regression adjustments were used, instead of considering the raw volume estimates, we estimated the standard deviation of the residuals of a simple linear regression regressing volume on time. This additional step gives a better indication of smoothness when the volume of the structure cannot be assumed to be constant over time, as is the case for those with AD. The table shows that PME yields the smoothest volume estimates for both hippocampi. This conclusion aligns with the visual evidence of inconsistency in the LPME estimates caused by the irregular shape of the hippocampus. However, when applied to the thalamus, LPME shows notable improvements in the smoothness of the associated volume estimates.

\begin{figure}[h]
  \centering
  %\includegraphics[width=\textwidth]{adni_plots/adni_cross_section}
  \includegraphics[height=8cm]{adni_plots/adni_cross_section}
  \caption{Left Hippocampus Cross Section. We can see that the LPME-approximated embedding function, shown in blue, routinely falls inside the boundaries of the surface in the observed data. In the fifth observation with $t=3.04$, the LPME-estimated surface has a gap in it. Gaps in the surface estimates frequently correspond with inaccurate volume estimates.}
  \label{fig:lhipp_cross_sections}
\end{figure}

\begin{figure}[h]
  \centering
  %\includegraphics[width=\textwidth]{adni_plots/adni_lthal_cross_section}
  \includegraphics[height=8cm]{adni_plots/adni_lthal_cross_section}
  \caption{Left Thalamus Cross Section. The cross sectional plots indicate that the PME and LPME algorithms fit much more closely to the regularly shaped thalamus surface.}
  \label{fig:lthal_cross_sections}
\end{figure}

@@

@@

@@

\begin{table}[ht]
  \centering
  \begin{tabular}{|c c c c c|}
    \hline
    Structure & Regression-Adjusted & Data & LPME & PME  \\
    \hline
    Left Hippocampus & No & 218 & 225 & \textbf{191} \\
    Right Hippocampus & No & 221 & 245 & \textbf{195} \\
    Left Hippocampus & Yes & 184 & 197 & \textbf{160} \\
    Right Hippocampus & Yes & 185 & 215 & \textbf{163} \\
    Left Thalamus & No & 82.9 & \textbf{49.7} & 72.0 \\
    Right Thalamus & No & 147 & \textbf{116} & 130 \\
    Left Thalamus & Yes & 68.3 & \textbf{30.7} & 57.9 \\
    Right Thalamus & Yes & 124 & \textbf{89.9} & 109 \\
    \hline
  \end{tabular}
  \caption{Mean of the empirical standard deviation of volume estimates computed for each ADNI participant. Entries that were not regression-adjusted include computations for raw volume estimates, while regression-adjusted entries compute the standard deviations of linear regression models regressing the volume estimates on time. }
  \label{table:adni_volume_sds}
\end{table}
